Distortion-corrected kernel density estimate on Riemannian manifolds

### 2022-10-11 paper/paper.Rmd

# Kernel density estimate on Riemannian manifolds - Tas {#kdermtas}

In this section we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm and the Metric Learning algorithm of @Perrault-Joncas2013-pq. We will do so by first introducing kernel density estimation for data in Euclidean space, then illustrating how this generalises to the estimator of @Pelletier2005-vu, when the data lie on some known manifold. Finally, we describe how our own proposed method generalises the estimator of @Pelletier2005-vu to the case where the underlying manifold is unknown. Denoting $\lambda$ as the Lebesgue measure of $\mathbb{R}^d$, letting $||.||$ be the usual Euclidean norm and following @Pelletier2005-vu we make these assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$ 
\begin{align}
\label{eq:kernelcondition}
& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{x}\|) \mathrm{d} \lambda(\pmb{x})=1;
(ii) \int_{\mathbb{R}^{d}} \pmb{x} K(\|\pmb{x}\|) \mathrm{d} \lambda(\pmb{x})=0;
(iii) \int_{\mathbb{R}^{d}}\|\pmb{x}\|^{2} K(\|\pmb{x}\|) \mathrm{d} \lambda(\pmb{x})<\infty; \\
& (iv) \supp K=[0; 1];
(v) \sup K(||\pmb{x}||)=K(0).
\end{align}
Note that these conditions are different (and in some cases stricter) that those normally used for kernel density estimation. For instance condition (iv) requires the support of the kernel to be bounded. The reasons for this will become clearer when we discuss the manifold setting in more detail. Also, for illustration purposes, in this section we pay particular attention to the uniform kernel for which $K(z)$ equals one if $0\leq z\leq1$ and zero otherwise. In our empirical section more general kernel functions can be, and are, employed.

Assume the data $\pmb{x}_i\in\mathbb{R}^d$ for $i=1,\dots,N$ and assuming a bandwidth matrix $r\bm{I}$ where $r$ is a global bandwidth, then the usual kernel density estimator at a point $\pmb{x}$ is given by

\begin{equation}
\label{eq:vkde}
\hat{f}(\pmb{x})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{||\pmb{x}-\pmb{x}_i||}{r}\right).
\end{equation}

The intuition behind this estimator is very clear for a uniform kernel. The density at a point $\pmb{x}$ is equal to the proportion of sample points that lie within a ball of radius $r$ centered at $\pmb{x}$, times a term that ensures the density integrates to 1. In general, the bandwidth matrix need not be proportional to the identity matrix. However, the intuition remains the same, only that the ball of radius $r$ centered at $\pmb{x}$ is found with respect to Mahalanobis distance rather than the usual Euclidean distance. For more on kernel density estimation in the Euclidean case see @Scott2015-vl and references therein.

@@

More formally, the volume density function can be explained as follows. Consider the exponential map around $\pmb{p}$, given by $exp_{\pmb{p}}(\pmb{q})$, mapping vectors in the tangent space $\pmb{v}\in T_{\pmb{p}}M$ to points the manifold $\pmb{q}\in M$ . Loosely, $\pmb{v}$ `points' in the direction of the geodesic between $\pmb{p}$ and $\pmb{q}$ and along this geodesic at uniform speed $||\pmb{v}||$ takes place in one unit of time. Now, consider a chart $\varphi$ mapping points in the neighbourhood of $\pmb{p}$, via the inverse of the exponential map, to these $\pmb{v}$ vectors, expressed in some local coordinate system. The volume density function is the square root of the determinant of the Riemannian metric expressed in this coordinate system. For more on the volume density function see @Le_Brigant2019-lj.

## Distortion corrected KDE

@@

```{r fivedcors, echo=FALSE, message=FALSE, eval=T}
load("figures/CorrelationTable_4d_N10000_4ml_radius10_r1.rda")
col.max <- apply(cors, 2, max)
max.match <- rep(col.max, each = 2) == cors
data.frame(cors) %>% 
  mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>% 
  kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
  # kable_styling(latex_options = "scale_down") %>% # automatically adjust the table to page width
  kable_paper(full_width = FALSE) %>%
  # column_spec(1, width = "6cm") %>% 
  column_spec(2, bold = max.match[,1]) %>%
  column_spec(3, bold = max.match[,2]) %>%
  column_spec(4, bold = max.match[,3]) %>%
  column_spec(5, bold = max.match[,4])
```
```{r fourdhdrtable, echo=FALSE, message=FALSE, eval=T}
load("figures/hdrtable_4d_N10000_4ml_radius10_r1.rda")
hdrtable %>% 
  kableExtra::kbl(caption = "Percentage comparison of correct highest density regions in density estimation of four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE, align = "c",
                  col.names = c(rep(c("DC-KDE", "KDE"), 4))
                  ) %>%
  # kable_classic("striped") %>% 
  kable_styling(latex_options = "scale_down") %>%
  kable_paper(full_width = FALSE) %>%
  # add_header_above(c("HDRs" = 1, rep(c("DC-KDE" = 1, "KDE" = 1), 4))) %>%
  add_header_above(c(" " = 1, "ISOMAP" = 2, "LLE" = 2, "Laplacian Eigenmaps" = 2, "UMAP" = 2)) %>% 
  # column_spec(1, bold = TRUE) %>% 
  column_spec(2, bold = hdrmax[,1]) %>% 
  column_spec(3, bold = hdrmax[,2]) %>% 
  column_spec(4, bold = hdrmax[,3]) %>% 
  column_spec(5, bold = hdrmax[,4]) %>% 
  column_spec(6, bold = hdrmax[,5]) %>% 
  column_spec(7, bold = hdrmax[,6]) %>% 
  column_spec(8, bold = hdrmax[,7]) %>% 
  column_spec(2*(1:4), border_left = T, background = grey(0.9))
```

>>>>

# Distortion Corrected Kernel density estimate on Riemannian manifolds - Tas {#kdermtas}

In this section we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding. Readers familiar with kernel density estimation may not be familiar with some of the nuances of manifolds. Therefore, before introducing our own estimator we first introduce kernel density estimation for data in Euclidean space, then illustrate in Section \@ref(Pellet) how this generalises to the estimator of @Pelletier2005-vu, when the data lie on some known manifold. In Section \@ref(MetLearn) we describe the Metric Learning algorithm of @Perrault-Joncas2013-pq, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates. By combining elements from the work of @Pelletier2005-vu and @Perrault-Joncas2013-pq we derive our own novel distortion corrected kernel density estimate in Section \@ref(DCKDE). This allows for kernel density estimation on unknown manifolds using any popular dimension reduction algorithm. To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance etc., but provide this information for readers less familiar with differential geometry in Appendix \@ref(riemgeo).

In the following we denote as $M$ the $d$-dimensional manifold from which our data are sampled. Points on this manifold are denoted $\pmb{p}$ in general with $\pmb{p}_1,\dots,\pmb{p}_n$ denoting the observed sample. Often $\pmb{p}_i$ will be high-dimensional vectors such that $\pmb{p}_i\in\mathbb{R}^m$ with $m>>d$. However this need not be that case, for instance $\pmb{p}_i$ may be probability distributions on a statistical manifold. The methods we propose for estimating the density at $\pmb{p}_i$ only require that there be available some sense of 'input' distance $d(\pmb{p}_i,\pmb{p}_j)$. We will apply dimension reduction algorithms to obtain an `output' embedding $\pmb{y}_1,\dots,\pmb{y}_n$ where $\pmb{y}_i\in\mathbb{R}^d$. Finally, we denote by $\lambda$ the Lebesgue measure of $\mathbb{R}^d$, letting $||.||$ be the usual Euclidean norm and following @Pelletier2005-vu we make these assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$ 
\begin{align}
\label{eq:kernelcondition}
& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=1;
(ii) \int_{\mathbb{R}^{d}} \pmb{y} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=0;
(iii) \int_{\mathbb{R}^{d}}\|\pmb{y}\|^{2} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})<\infty; \\
& (iv) \supp K=[0; 1];
(v) \sup K(||\pmb{y}||)=K(0).
\end{align}
Note that these conditions are different (and in some cases stricter) that those normally used for kernel density estimation. For instance condition (iv) requires the support of the kernel to be bounded. The reasons for this will become clearer when we discuss the manifold setting in more detail. Also, for illustration purposes, in this section we pay particular attention to the uniform kernel for which $K(z)$ equals one if $0\leq z\leq1$ and zero otherwise. In our empirical section more general kernel functions can be, and are, employed.

For data $\pmb{y}_i\in\mathbb{R}^d$ for $i=1,\dots,N$ and assuming a bandwidth matrix $r\bm{I}$ where $r$ is a global bandwidth, then the usual kernel density estimator at a point $\pmb{y}$ is given by

\begin{equation}
\label{eq:vkde}
\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{||\pmb{y}-\pmb{y}_i||}{r}\right).
\end{equation}

The intuition behind this estimator is very clear for a uniform kernel. The density at a point $\pmb{y}$ is equal to the proportion of sample points that lie within a ball of radius $r$ centered at $\pmb{y}$, times a term that ensures the density integrates to 1. In general, the bandwidth matrix need not be proportional to the identity matrix. However, the intuition remains the same, only that the ball of radius $r$ centered at $\pmb{y}$ is found with respect to Mahalanobis distance rather than the usual Euclidean distance. For more on kernel density estimation in the Euclidean case see @Scott2015-vl and references therein.

While one could in principle use a standard kernel density on the output from a dimension reduction algorithm, it must be noted that the density of the output vectors $\pmb{y}_i$ is different to the density on the manifold itself. As a non-linear transformation, any dimension reduction algorithm `distorts' the density. Furthermore any analysis based on density estimates of the output embedding will be dependent on the choice of dimension reduction algorithm since each different algorithm will distort the density in a different way.

## Kernel Density estimation on manifolds {#Pellet}

@@

More formally, the volume density function can be explained as follows. Consider the exponential map around $\pmb{p}$, given by $exp_{\pmb{p}}(\pmb{q})$, mapping vectors in the tangent space $\pmb{v}\in T_{\pmb{p}}M$ to points the manifold $\pmb{q}\in M$ . Loosely, $\pmb{v}$ `points' in the direction of the geodesic between $\pmb{p}$ and $\pmb{q}$ and travel along this geodesic at uniform speed $||\pmb{v}||$ takes place in one unit of time. Now, consider a chart $\varphi$ mapping points in the neighbourhood of $\pmb{p}$, via the inverse of the exponential map, to these $\pmb{v}$ vectors, expressed in some local coordinate system. The volume density function is the square root of the determinant of the Riemannian metric expressed in this coordinate system. For more on the volume density function see @Le_Brigant2019-lj.

## Riemannian metric estimation {#MetLearn}

To be able to apply the estimator of @Pelletier2005-vu to the case there the manifold is not known, but where an output embedding from a dimension reduction algorithm is available requires additional estimates of the Riemannian metric. Formally, the Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $\langle,\rangle_g$ on the tangent space $T_{\pmb{p}} M$ for every point $\pmb{p} \in M$. The inner product between two tangent vectors $u,v \in T_{\pmb{p}}M$, given by $\langle u,v \rangle_g$[^1], can be used to define geometric quantities for example angles on a manifold are given by $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$, while distances and volumes on manifolds are also defined with reference to the Riemannian metric. While tangent vectors, the Riemannian metric and the geometric quantities they define are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinates systems. This is precisely the situation when data on a manifold are mapped to $d$-dimensional Euclidean vectors $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ via a dimension reduction algorithm. After this mapping, angles, distances and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortion. To alleviate this issue @Perrault-Joncas2013-pq propose a method to augment $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ with $d\times d$ positive definite matrices $\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n$ at each data point.  These matrices estimate the Reimannian metric in local coordinates defined by the dimension reduction algorithm, for example the angle between $\pmb{p}_j$ and $\pmb{p}_k$ at $\pmb{p}_i$ depends on the inner product $(\pmb{y}_j-\pmb{y}_i)'\pmb{H}(\pmb{y}_k-\pmb{y}_i)$ rather than the usual Euclidean inner product $(\pmb{y}_j-\pmb{y}_i)'\pmb{H}(\pmb{y}_k-\pmb{y}_i)$.

Locally isometric embedding to go here...

While full details are provided in @Perrault-Joncas2013-pq , we briefly describe the Learn Metric algorithm here. There are four main steps in the algorithm. First a weighted neighborhood graph is constructed, with edges between $\pmb{p}_i$ and $\pmb{p}_j$ when $\pmb{p}_i$ is a K-nearest neighbor of $\pmb{p}_j$ or vice versa, and edge weights depending on the distance between $\pmb{p}_i$ and $\pmb{p}_j$ on the manifold. Second, the discrete Laplacian on this graph $\hat{\mathcal{L}}_{\varepsilon,n}$ is estimated [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph. Third, a dimension reduction method is applied to obtain the output embedding $\pmb{y}_1,\dots,\pmb{y}_n$. Fourth, the Riemannian metric at each point is estimated by exploiting the connection between the Riemannian metric and the Laplace Beltrami operator (to which the graph Laplacian at step 2 is a discrete estimator). Full details on these four steps are provided in Algorithm \@ref(alg:learnmetric). This algorithm is implemented in a Python library *megaman* [@McQueen2016-xz] although our own results are based on a re-implementation of the algorithm in *R*.

## Distortion corrected KDE {#DCKDE}

@@

```{r fivedcors, echo=FALSE, message=FALSE, eval=T}
load("figures/CorrelationTable_4d_N10000_4ml_radius10_r1.rda")
col.max <- apply(cors, 2, max)
max.match <- rep(col.max, each = 2) == cors
data.frame(cors) %>% 
  mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>% 
  kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
  # kable_styling(latex_options = "scale_down") %>% # automatically adjust the table to page width
  kable_paper(full_width = FALSE) %>%
  # column_spec(1, width = "6cm") %>% 
  column_spec(2, bold = max.match[,1]) %>%
  column_spec(3, bold = max.match[,2]) %>%
  column_spec(4, bold = max.match[,3]) %>%
  column_spec(5, bold = max.match[,4])
```

```{r fourdhdrtable, echo=FALSE, message=FALSE, eval=F}
load("figures/HDRTable_4d_N10000_4ml_radius10_r1.rda")
hdrtable %>% 
  kableExtra::kbl(caption = "Percentage comparison of correct highest density regions in density estimation of four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE, align = "c",
                  col.names = c(rep(c("DC-KDE", "KDE"), 4))
                  ) %>%
  # kable_classic("striped") %>% 
  kable_styling(latex_options = "scale_down") %>%
  kable_paper(full_width = FALSE) %>%
  # add_header_above(c("HDRs" = 1, rep(c("DC-KDE" = 1, "KDE" = 1), 4))) %>%
  add_header_above(c(" " = 1, "ISOMAP" = 2, "LLE" = 2, "Laplacian Eigenmaps" = 2, "UMAP" = 2)) %>% 
  # column_spec(1, bold = TRUE) %>% 
  column_spec(2, bold = hdrmax[,1]) %>% 
  column_spec(3, bold = hdrmax[,2]) %>% 
  column_spec(4, bold = hdrmax[,3]) %>% 
  column_spec(5, bold = hdrmax[,4]) %>% 
  column_spec(6, bold = hdrmax[,5]) %>% 
  column_spec(7, bold = hdrmax[,6]) %>% 
  column_spec(8, bold = hdrmax[,7]) %>% 
  column_spec(2*(1:4), border_left = T, background = grey(0.9))
```
